Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-05-31 paper/main.tex
implement plots

\paragraph{TODO}
\begin{itemize}
    \item Change all parameter-dependent matrices to $\mtx{B}(t)$ to avoid confusion with $\mtx{A}$
    \item Change citation style to \textbf{[1]}, \textbf{[2]}, \dots
    \item Re-run numerical experiments with simplified and improved implementation
\end{itemize}
}

@@

In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}. The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive. Therefore, many techniques have been proposed for approximating $\phi$. For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smooth spectral density
\begin{equation}
    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t \mtx{I}_n - \mtx{A}))
\end{equation}
for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. This translates the problem to computing the trace of the parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$. Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive. Therefore, we estimate its trace implicitly by approximating products of the matrix function with vectors. For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}. Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation}, but they have not yet been analyzed. A related technique is dynamic trace estimation for sequences $\mtx{B}_1, \dots, \mtx{B}_m$ of similar matrices \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}, which achieves guarantees for the estimate $b_i$ of $\Trace(\mtx{B}_i)$ of the form $\mathbb{P}(|\Trace(\mtx{B}_i) - b_i| \geq \varepsilon) \leq \delta$ for $i=1,\dots, m$. A drawback of this type of bound is that when we aim for $\varepsilon$ accuracy for all $i=1,\dots, m$ simultaneously, the probability will grow linearly with $m$ by the union bound. Therefore, instead of discretizing $[a, b]$ with $t_1, \ldots, t_m$ and looking at the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way to quantify the error is to use $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate for each $t \in [a, b]$.

@@

The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly. For $n_{\mtx{\Psi}} = 0$ we end up with \refequ{equ:nystrom-trace-estimator}.
%Chebyshev $\mathcal{O}(n_t (m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + n_{\mtx{\Omega}}^3) + mn^2(n_{\mtx{\Psi}} + n_{\mtx{\Omega}}))$.

@@

\todo{Explain short-circuit mechanism -> may not be too crucial for directly calculating pseudo-inverse}

@@

% Acknowledge advice from David

>>>>

\paragraph{TODO}
\begin{itemize}
    \item Change all parameter-dependent matrices to $\mtx{B}(t)$ to avoid confusion with $\mtx{A}$
    \item Change citation style to \textbf{[1]}, \textbf{[2]}, \dots
    \item Re-run numerical experiments with simplified and improved implementation (test pinvh against lstsq)
\end{itemize}
}

@@

In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}. The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive. Therefore, many techniques have been proposed for approximating $\phi$. For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smooth spectral density
\begin{equation}
    \phi_{\sigma}(t) = \frac{1}{n} \sum_{i=1}^{n} g_{\sigma}(t - \lambda_i) = \frac{1}{n} \Trace(g_{\sigma}(t \mtx{I}_n - \mtx{A}))
\end{equation}
for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. This translates the problem to computing the trace of a continuous parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$. Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive. Therefore, we estimate its trace implicitly by approximating products of the matrix function with vectors. For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}. Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation}, but they have not yet been analyzed. A related technique is dynamic trace estimation for sequences $\mtx{B}_1, \dots, \mtx{B}_m$ of similar matrices \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}, which achieves guarantees for the estimate $b_i$ of $\Trace(\mtx{B}_i)$ of the form $\mathbb{P}(|\Trace(\mtx{B}_i) - b_i| \geq \varepsilon) \leq \delta$ for $i=1,\dots, m$. A drawback of this type of bound is that when we aim for $\varepsilon$ accuracy for all $i=1,\dots, m$ simultaneously, the probability will grow linearly with $m$ by the union bound. Therefore, instead of discretizing $[a, b]$ with $t_1, \ldots, t_m$ and looking at the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way to quantify the error is to use $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate for each $t \in [a, b]$.

@@

The algorithm also accomodates the two other estimators introduced in \refsec{sec:analysis}: For $n_{\mtx{\Omega}} = 0$ we recover \refequ{equ:hutchinson-trace-estimator}, and the complexity for each evaluation of the estimate decreases significantly. For $n_{\mtx{\Psi}} = 0$ we end up with \refequ{equ:nystrom-trace-estimator}.

\todo{Clearly need not divide by $n_{\mtx{\Psi}}=0$, hence just ignore this term. }
%Chebyshev $\mathcal{O}(n_t (m \log(m) + m (n_{\mtx{\Omega}}^2 + n_{\mtx{\Omega}} n_{\mtx{\Psi}} + 1) + n_{\mtx{\Omega}}^3) + mn^2(n_{\mtx{\Psi}} + n_{\mtx{\Omega}}))$.

@@

\todo{Explain short-circuit mechanism -> very crucial if matrix can be zero}

@@

% Acknowledge advice from David and Data/Parameters from Lin Lin
